A Process for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE)

### 2025-02-12 manuscript/manuscript.qmd
added pagebreak
implemented GAH comments

**Authors**: Janick Weberpals^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, Pallavi Mishra-Kalyani^3^, Georg Hahn^1^, Priyanka Anand^1^, ..., Sebastian Schneeweiss^1^, Shirley V. Wang^1^

@@

@@

Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products. With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology where potential use cases include the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures..

@@

The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project [@encoreFDA] aims to extend this work to the field of oncology. Studies in oncology come with their own unique set of challenges which must be systematically explored and understood. Building on a process co-developed with the FDA through RCT DUPLICATE[@wang2023emulation], this expansion to oncology will emulate 12 randomized clinical trials using multiple specialty oncology EHR data sources. The process will emphasize transparency and include documented data fitness assessment[@rivera2024oncology; @gatto2022structured] for each RWD source with respect to each trial emulation as well as extensive sensitivity analyses to assess robustness of findings.

@@

The ENCORE project will use data from a four oncology-specific electronic health records (EHR)-derived specialty oncology data sources: ConcertAI, COTA, Flatiron Health, McKesson/Ontada. All available databases draw from a comprehensive national sample of patients with cancer in the US with detailed EHR-derived information on the information necessary to study medication effectiveness in oncology. A detailed description and methodology on how patients are sampled will be provided with each trial emulation protocol. For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results. If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed and random effects models.

@@

According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], the next step will outline tables that describe how eligibility criteria will be ascertained using a color-coded heatmap that will indicate the level of confidence on how well each criterion can be emulated in each selected database. As there are general eligibility criteria in oncology trials which either won't be possible to emulate (e.g., physician-assessed survival prognosis of xy months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.

@@

he distribution of patient characteristics, stratified by exposure status, will be examined in Table 1's before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT. Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, eposure cohorts are conditionally exchangeable at baseline and resulting sample sizes are still sufficient after matching or weighting. At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design and analytic choices based on known outcome information.

@@

Causal analyses of observational data may not have the same pre-requisites in terms of formal hypothesis testing and statistical power than RCTs since the number of 'recruited' patients is given and cannot be influenced.[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreements metrics between RCT and RWD results. Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients. To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power $\beta$ will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]

@@

where $\hat{\theta}$ are the treatment effect estimates (hazard ratios or median survival times) and $\text{Var}(\hat{\theta})$ are the corresponding variances. The resulting SMDs will be interpreted such that with an SMD of 1.00, the effect estimate from the RCT and the RWE emulation are 1 standard deviation apart. For an $\alpha$-level of 0.05, the null hypothesis of no difference would be rejected whenever $|Z| > 1.96$.

@@

Similarly, survival probabilities for individual time points will be estimated in each imputed and propensity score matched or weighted dataset according the Kaplan-Meier method.[@kaplan1958nonparametric] Since survival probabilities typically don't follow normal distributions which are required to apply Rubin's rule, these will be transformed through a complementary log-log transformation $log(-log(1-pr(surv)))$ with $pr(surv)$ denoting the survival probability at a given time during follow-up.[@warwick28685; @morisot2015prostate] The transformed survival probabilities are then pooled across imputed datasets and individual time points following Rubin's rule and back-transformed via $1-exp(-exp(qbar))$ with $qbar$ denoting the pooled survival probability. The median survival time can be finally determined by extracting the time point during follow-up at which the survival probability drops below 0.5 for the first time.

@@

To ensure a transparent, reproducible and consistent way of deriving analytic cohorts and performing statistical analyses across trial emulations, we developed and internal R package `encore.io` with parameterized functions. Detailed documentation can be found in the Supplementary Material.

@@

Building off an established approach based on the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using multiple EHR data sources to inform the potential use of RWE for regulatory science in oncology. The project focuses on four common cancers and will evaluate the agreement of treatment effect estimates between RCTs and their respective emulations. Historically, administrative health claims databases have been the backbone of most research in RWE. With increasing access to EHR and a maturing set of methodological approaches to draw causal inferences from such data, the ENCORE project is one of the first of its kind to evaluate when and how RWD can be used to deliver similar causal conclusions compared to RCTs in the field of oncology.

The RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized] identified multiple emulation challenges which we expect to also encounter in ENCORE. One particular aspect that we may not always be able emulate is the exact distribution of patient characteristics of the trial population. This can be due to the lack of data granularity and comprehensiveness to emulate relevant eligibility criteria or the fact that large pivotal trials in oncology are typically conducted in multiple countries worldwide, which we won't be able to mirror given that all considered databases reflect the US only. This may be a critical factor especially given that the pathophysiology, prognosis and factors that drive heterogeneous treatment outcomes of certain cancers differ between countries. For example, some cancer indications (e.g., GI cancers) or some genetic mutations (e.g., EGFR mutations) are much more prevalent in Asian countries compared to the US or Europe.

@@

```{r}
#| label: fig-balance
#| fig-cap: "Assessment of a) covariate bbalance and b) distributional balance of a prognostic score for overall survival before and after propensity score matching or weighting across multiple imputed datasets."
#| fig-widh: 4
#| fig-height: 7

>>>>

**Authors**: Janick Weberpals^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, Pallavi Mishra-Kalyani^3^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^4^, Janet Espirito^5^, Nicholas J Robert^5^, Andrew J. Belli^6^, Sebastian Schneeweiss^1^, Shirley V. Wang^1^

@@

^4^ ConcertAI, Cambridge, MA, USA

^5^ Ontada, Boston, MA, USA

^6^ COTA, Inc., New York, NY, USA

@@

Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products. With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology where potential use cases include the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.

@@

The *Emulation of Comparative Oncology Trials with Real-world Evidence* (ENCORE) project [@encoreFDA] aims to extend this work to the field of oncology. Studies in oncology come with their own unique set of challenges which must be systematically explored and understood. Building on a process co-developed with the FDA through RCT DUPLICATE[@wang2023emulation], this expansion to oncology will emulate 12 randomized clinical trials using multiple specialty oncology EHR data sources. The process will emphasize transparency and include documented data fitness assessment[@rivera2024oncology; @gatto2022structured] for each RWD source with respect to each trial emulation as well as extensive sensitivity analyses to assess robustness of findings.

@@

The ENCORE project will use data from a four oncology-specific electronic health records (EHR)-derived specialty oncology data sources: ConcertAI, COTA, Flatiron Health, McKesson/Ontada. All available databases draw from a comprehensive national sample of patients with cancer in the US with detailed EHR-derived information on the information necessary to study medication effectiveness in oncology. A detailed description and methodology on how patients are sampled will be provided with each trial emulation protocol. For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results. If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects (weights reflect the inverse of the variance of each database study) and random effects (weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) models.[@nikolakopoulou2014interpret]

@@

According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], the next step will outline tables that describe how eligibility criteria will be ascertained using a color-coded heatmap that will indicate the level of confidence on how well each criterion can be emulated in each selected database. As there are general eligibility criteria in oncology trials which either will not be possible to emulate (e.g., physician-assessed survival prognosis of xy months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.

@@

The distribution of patient characteristics, stratified by exposure status, will be examined in Table 1's before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT. Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, exposure cohorts are conditionally exchangeable at baseline and resulting sample sizes are still sufficient after matching or weighting. At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design and analytic choices based on known outcome information.

@@

Causal analyses of observational data may not have the same pre-requisites in terms of formal hypothesis testing and statistical power than RCTs since the number of 'recruited' patients is given and cannot be influenced.[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreements metrics between RCT and RWD results. Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients. To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power (1-$\beta$) will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]

@@

where $\hat{\theta}_{RCT}$ and $\hat{\theta}_{RWE}$ are the treatment effect estimates (hazard ratios or median survival times) and $\text{Var}(\hat{\theta}_{RCT})$ and $\text{Var}(\hat{\theta}_{RWE})$ are the corresponding variances for RCT and RWE, respectively. The resulting SMDs will be interpreted such that with an SMD of 1.00, the effect estimate from the RCT and the RWE emulation are 1 standard deviation apart. For an $\alpha$-level of 0.05, the null hypothesis of no difference would be rejected whenever $|Z| > 1.96$.

@@

Similarly, survival probabilities for individual time points will be estimated in each imputed and propensity score matched or weighted dataset according the Kaplan-Meier method.[@kaplan1958nonparametric] Since survival probabilities typically do not follow normal distributions which are required to apply Rubin's rule, these will be transformed through a complementary log-log transformation $log(-log(1-pr(surv)))$ with $pr(surv)$ denoting the survival probability at a given time during follow-up.[@warwick28685; @morisot2015prostate] The transformed survival probabilities are then pooled across imputed datasets and individual time points following Rubin's rule and back-transformed via $1-exp(-exp(qbar))$ with $qbar$ denoting the pooled survival probability. The median survival time can be finally determined by extracting the time point during follow-up at which the survival probability drops below 0.5 for the first time.

@@

To ensure a transparent, reproducible and consistent way of deriving analytic cohorts and performing statistical analyses across trial emulations, we developed an internal R package `encore.io` with parameterized functions. Detailed documentation can be found in the Supplementary Material.

@@

Building on an established approach based on the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using multiple EHR data sources to inform the potential use of RWE for regulatory science in oncology. The project focuses on four common cancers and will evaluate the agreement of treatment effect estimates between RCTs and their respective emulations. Historically, administrative health claims databases have been the backbone of most research in RWE. With increasing access to EHR and a maturing set of methodological approaches to draw causal inferences from such data, the ENCORE project is one of the first of its kind to evaluate when and how RWD can be used to deliver similar causal conclusions compared to RCTs in the field of oncology.

The RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized] identified multiple emulation challenges which we expect to also encounter in ENCORE. One particular aspect that we may not always be able emulate is the exact distribution of patient characteristics of the trial population. This can be due to the lack of data granularity and comprehensiveness to emulate relevant eligibility criteria or the fact that large pivotal trials in oncology are typically conducted in multiple countries worldwide, which we will not be able to mirror given that all considered databases reflect the US only. This may be a critical factor especially given that the pathophysiology, prognosis and factors that drive heterogeneous treatment outcomes of certain cancers differ between countries. For example, some cancer indications (e.g., GI cancers) or some genetic mutations (e.g., EGFR mutations) are much more prevalent in Asian countries compared to the US or Europe.

@@

```{r}
#| label: fig-balance
#| fig-cap: "Assessment of a) covariate balance and b) distributional balance of a prognostic score for overall survival before and after propensity score matching or weighting across multiple imputed datasets."
#| fig-widh: 4
#| fig-height: 7
